---
title: "lab11"
author: "Meha Thakur PID 16020450"
format: pdf
---

## Background

We saw last tday that PDB has 209886 entries. UniprotKB has 199579901 entries. PDB has only 0.1% coverage of the main sequence database. 

Enter AlphaFold data base (AFDB) - a site hosting computed models that attempts to provide computed models for all sequences in Uniprot. https://alphafold.ebi.ac.uk/



```{r}

```


## AlphaFold

AlphaFold has 3 main outputs

- predicted coordinates
- local quality score **pLDDT** (one for each amino acid)
- Second quality score **PAE** - Predicted Aligned Error (for each aa)

We can run ALphaFold ourselves if we're not happy with AFDB.

## Interpreting/analysing AF results in R

```{r}
results_dir <- "HIVPR_dimer_23119"

# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)

```

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```
```{r}
library(bio3d)
library(msa)

pdbs

```

```{r}
rd <- rmsd(pdbs, fit=T)

range(rd)
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```

plotting
```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")

```

